---
title: "Independent Bayesian Validation of the Np Two-Acid GLS Age Calculator"
output:
  html_document:
    toc: true
    toc_float: true
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
library(ggplot2)
library(dplyr)

source("R/mcmc.R")
source("R/age_calibration.R")
source("R/age_inversion.R")
source("R/gls_calculator.R")
```

## Overview

This notebook supports section 4.1 ("Validation by independent Bayesian
method") of the Np AAR age-calculator manuscript. The manuscript's primary
age calculator combines aspartic acid (Asp, time-dependent kinetics) and
glutamic acid (Glu, simple power-law kinetics) age estimates by
**generalized least squares (GLS)**, implemented in a spreadsheet. Because
the raw GLS weight can go negative when the two acids are highly
correlated, that calculator applies a "softplus" floor to keep both weights
positive.

As an independent check, this notebook performs a **Bayesian MCMC
inversion** of the same two-acid evidence: given the manuscript's own
fitted TDK (Asp) and SPK (Glu) calibration curves and their correlated
residual structure (`R/gls_calculator.R`, a faithful re-implementation of
the spreadsheet's `Constants` sheet), it infers a full posterior
distribution for a sample's age from its Asp and Glu D/L using a bivariate
normal likelihood (`log_lik_two_acid_age()` in `R/age_inversion.R`), rather
than an algebraic GLS combination of two point estimates. Because this is
an ordinary Bayesian posterior, the two acids combine coherently for any
value of the residual correlation with no floor or smoothing constant
needed.

This validation isolates the **inversion/combination** step specifically:
it starts from the manuscript's own published, fixed calibration curves
(no calibration constants are re-estimated here), so any agreement (or
disagreement) below reflects how the two acids' evidence is combined, not
differences in how the underlying curves were fit. The Bayesian approach
is used here only as an independent test of the spreadsheet calculator,
not as a proposed replacement calibration.

## Calibration models

Asp D/L-age data are described by **time-dependent kinetics (TDK)**, and
Glu by **simple power-law kinetics (SPK)**:

$$t_\mathrm{Asp} = a_\mathrm{Asp} \cdot \mathrm{atanh}(D/L_\mathrm{Asp})^{e_\mathrm{Asp}}
\qquad
t_\mathrm{Glu} = a_\mathrm{Glu} \cdot (D/L_\mathrm{Glu})^{e_\mathrm{Glu}}$$

`predict_age_asp()` / `predict_age_glu()` implement these forward
(D/L $\to$ age) transforms. The scale/exponent constants and residual
correlation (`gls_calibration_constants` in `R/gls_calculator.R`) are taken
directly from the manuscript's own OLS fit of the 128-sample calibration
set, not re-estimated here.

## The calibration dataset

The 128-sample Np calibration dataset (Arctic Ocean, Nordic Seas, North
Atlantic, and Southern Ocean cores; ages 2 ka to 1.4 Myr).

```{r load_data}
np <- read.csv("data/np_calibration.csv")

# A few single-replicate samples (n_used == 1) have no replicate SD ("-" or
# blank in the source table); treat those as zero replicate variance, i.e.
# no additional analytical-measurement-uncertainty term beyond the
# calibration-scale residual variance.
np$SD_Asp <- suppressWarnings(as.numeric(np$SD_Asp))
np$SD_Asp[is.na(np$SD_Asp)] <- 0
np$SD_Glu <- suppressWarnings(as.numeric(np$SD_Glu))
np$SD_Glu[is.na(np$SD_Glu)] <- 0

cal_data <- data.frame(age_ka = np$age_ka, DL_Asp = np$DL_Asp, DL_Glu = np$DL_Glu)
nrow(cal_data)
head(cal_data)
```

```{r plot_data, fig.height=4}
ggplot(np, aes(x = DL_Asp, y = DL_Glu, colour = age_ka)) +
  geom_point(size = 1.8) +
  scale_colour_viridis_c(trans = "log10", name = "Age (ka)") +
  labs(x = "D/L Asp", y = "D/L Glu",
       title = "Np calibration dataset: Asp vs. Glu D/L") +
  theme_bw()
```

## The published calibration constants

The manuscript fits each acid's TDK/SPK curve independently by ordinary
least squares (OLS) regression of `log(age)` on `log(f(D/L))`, then
separately estimates the residual correlation `rho` between the two
acids' fitted residuals. `gls_calibration_constants` (`R/gls_calculator.R`)
reproduces that fit exactly:

```{r ols_check}
fit_asp <- lm(log(age_ka) ~ log(atanh(DL_Asp)), data = np)
fit_glu <- lm(log(age_ka) ~ log(DL_Glu), data = np)
data.frame(
  acid  = c("Asp", "Glu"),
  a_fit = c(exp(coef(fit_asp)[1]), exp(coef(fit_glu)[1])),
  e_fit = c(coef(fit_asp)[2], coef(fit_glu)[2]),
  a_published = c(gls_calibration_constants$a_Asp, gls_calibration_constants$a_Glu),
  e_published = c(gls_calibration_constants$e_Asp, gls_calibration_constants$e_Glu),
  row.names = NULL
)
```

Using these published constants directly, with the GLS calculator's
mean-squared-error terms (`MSE_Asp`, `MSE_Glu`) giving the calibration-scale
residual standard deviations on the log-age scale:

```{r calib_params}
calib_params <- list(
  a_Asp     = gls_calibration_constants$a_Asp,
  e_Asp     = gls_calibration_constants$e_Asp,
  sigma_Asp = sqrt(gls_calibration_constants$MSE_Asp),
  a_Glu     = gls_calibration_constants$a_Glu,
  e_Glu     = gls_calibration_constants$e_Glu,
  sigma_Glu = sqrt(gls_calibration_constants$MSE_Glu),
  rho       = gls_calibration_constants$rho
)
calib_params
```

```{r plot_fit, fig.height=4.5}
dl_seq <- seq(0.01, 0.99, length.out = 200)

curves <- rbind(
  data.frame(acid = "Aspartic acid (TDK)", DL = dl_seq,
             age_ka = predict_age_asp(dl_seq, calib_params$a_Asp, calib_params$e_Asp)),
  data.frame(acid = "Glutamic acid (SPK)", DL = dl_seq,
             age_ka = predict_age_glu(dl_seq, calib_params$a_Glu, calib_params$e_Glu))
)

obs <- rbind(
  data.frame(acid = "Aspartic acid (TDK)", DL = np$DL_Asp, age_ka = np$age_ka),
  data.frame(acid = "Glutamic acid (SPK)", DL = np$DL_Glu, age_ka = np$age_ka)
)

ggplot() +
  geom_point(data = obs, aes(x = DL, y = age_ka), size = 1.2, alpha = 0.4) +
  geom_line(data = curves, aes(x = DL, y = age_ka), colour = "steelblue", linewidth = 1) +
  scale_y_log10() +
  facet_wrap(~acid, scales = "free_x") +
  labs(x = "D/L", y = "Age (ka BP)",
       title = "Np AAR age calibration curves: published OLS fit") +
  theme_bw()
```

## Two worked examples

Two worked examples from the manuscript's spreadsheet calculator provide a
direct check: their GLS-combined ages and 66% ("likely") prediction
intervals can be compared against the Bayesian posterior for the same D/L
inputs, using the published calibration constants above (i.e. isolating
the inversion step).

```{r stage2a_examples, cache=TRUE}
examples <- list(
  "Example 1" = list(DL_Asp = 0.350, DL_Glu = 0.153,
                     gls_median = 176.25, gls_lo66 = 119.47, gls_hi66 = 260.03),
  "Example 2" = list(DL_Asp = 0.272, DL_Glu = 0.110,
                     gls_median = 82.59, gls_lo66 = 49.97, gls_hi66 = 136.52)
)

set.seed(1)
mcmc_examples <- lapply(examples, function(ex) {
  invert_two_acid_age(DL_Asp_obs = ex$DL_Asp, DL_Glu_obs = ex$DL_Glu,
                      calib_params = calib_params, n_iter = 30000)
})

comparison <- do.call(rbind, lapply(names(examples), function(nm) {
  post <- summarize_age_posterior(mcmc_examples[[nm]], burnin = 5000)
  data.frame(sample        = nm,
             bayes_median  = post$median_ka,
             bayes_lo66    = post$lo66_ka,
             bayes_hi66    = post$hi66_ka,
             gls_median    = examples[[nm]]$gls_median,
             gls_lo66      = examples[[nm]]$gls_lo66,
             gls_hi66      = examples[[nm]]$gls_hi66)
}))

knitr::kable(comparison, digits = 1)
```

```{r plot_examples, fig.height=4}
plot_age_posterior(mcmc_examples[["Example 1"]], burnin = 5000,
                   compare_ages_ka = c("GLS combined" = 176.25)) +
  labs(subtitle = "Example 1: DL_Asp = 0.350, DL_Glu = 0.153")
```

## Full calibration-set comparison

The real validation in the manuscript compares GLS against a Bayesian
inversion across the **entire 128-sample calibration set**, not just two
worked examples. For every sample we compute:

- the reference **GLS** age and 66% interval via `predict_age_gls()`, using
  the manuscript's own fitted calibration constants
  (`gls_calibration_constants`) and each sample's observed D/L, replicate
  SD, and number of replicates (`n_used`);
- the **Bayesian** posterior median age and 66% credible interval via
  `invert_two_acid_age()`, using the same published calibration constants.

```{r stage2b_full, cache=TRUE}
n_iter_full  <- 6000
burnin_full  <- 1500

run_full_comparison <- function(calib_params, seed) {
  set.seed(seed)
  res <- lapply(seq_len(nrow(np)), function(i) {
    row <- np[i, ]

    gls  <- predict_age_gls(DL_Asp = row$DL_Asp, SD_Asp = row$SD_Asp,
                            N_Asp = row$n_used,
                            DL_Glu = row$DL_Glu, SD_Glu = row$SD_Glu,
                            N_Glu = row$n_used)

    mcmc_i <- invert_two_acid_age(DL_Asp_obs = row$DL_Asp, DL_Glu_obs = row$DL_Glu,
                                  calib_params = calib_params, n_iter = n_iter_full)
    bayes  <- summarize_age_posterior(mcmc_i, burnin = burnin_full)

    data.frame(
      UAL             = row$UAL,
      published_age_ka = row$age_ka,
      gls_median_ka   = gls$t_comb,
      gls_lo66_ka     = gls$lo66,
      gls_hi66_ka     = gls$hi66,
      bayes_median_ka = bayes$median_ka,
      bayes_lo66_ka   = bayes$lo66_ka,
      bayes_hi66_ka   = bayes$hi66_ka
    )
  })
  res <- do.call(rbind, res)
  res |>
    mutate(
      pct_diff_median  = 100 * abs(bayes_median_ka - gls_median_ka) / gls_median_ka,
      gls_width66      = gls_hi66_ka - gls_lo66_ka,
      bayes_width66    = bayes_hi66_ka - bayes_lo66_ka,
      pct_diff_width66 = 100 * abs(bayes_width66 - gls_width66) / gls_width66
    )
}

full_results <- run_full_comparison(calib_params, seed = 7)

summary_table <- data.frame(
  median_pct_diff_age = median(full_results$pct_diff_median),
  max_pct_diff_age    = max(full_results$pct_diff_median),
  median_width_ratio  = median(full_results$bayes_width66 / full_results$gls_width66)
)
knitr::kable(summary_table, digits = 2,
             col.names = c("Median abs. % diff (age)",
                           "Max abs. % diff (age)", "Median width ratio (Bayes/GLS)"))
```

```{r plot_parity, fig.height=5}
ggplot(full_results, aes(x = gls_median_ka, y = bayes_median_ka)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.6, colour = "steelblue") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "GLS combined age (ka, spreadsheet method)",
       y = "Bayesian posterior median age (ka)",
       title = "GLS vs. Bayesian age estimates across the Np calibration set",
       subtitle = "Dashed line: 1:1") +
  theme_bw()
```

```{r plot_interval_width, fig.height=5}
ggplot(full_results, aes(x = gls_width66, y = bayes_width66)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.6, colour = "steelblue") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "GLS 66% interval width (ka)",
       y = "Bayesian 66% credible interval width (ka)",
       title = "Prediction/credible interval widths: GLS vs. Bayesian",
       subtitle = "Dashed line: 1:1") +
  theme_bw()
```

```{r table_full, echo=FALSE}
knitr::kable(head(full_results, 10), digits = 2,
             caption = "First 10 of 128 calibration samples (full table omitted for brevity)")
```

**Point estimates:** the Bayesian inversion agrees with GLS closely, a
median absolute difference of a few percent, driven almost entirely by the
handful of samples where the GLS softplus regularisation nudges the raw
(possibly negative) GLS weight away from its unconstrained value, a step
the Bayesian posterior does not need.

**Uncertainty (interval widths):** the Bayesian 66% credible interval
widths track the GLS 66% interval widths closely (ratio near 1) across the
calibration set.

## Comparison against independently known ages

Both methods (GLS spreadsheet and Bayesian inversion) share the same
published calibration curves, so both can be checked directly against the
calibration set's own independently established ages (`age_ka`, from
radiocarbon, oxygen-isotope stratigraphy, etc., depending on the source
study) for all 128 samples.

```{r stage3_known_ages}
two_way <- rbind(
  data.frame(UAL = full_results$UAL,
             known_age_ka = full_results$published_age_ka,
             method   = "Spreadsheet (GLS)",
             est_age_ka = full_results$gls_median_ka,
             lo66 = full_results$gls_lo66_ka,
             hi66 = full_results$gls_hi66_ka),
  data.frame(UAL = full_results$UAL,
             known_age_ka = full_results$published_age_ka,
             method   = "Bayesian inversion",
             est_age_ka = full_results$bayes_median_ka,
             lo66 = full_results$bayes_lo66_ka,
             hi66 = full_results$bayes_hi66_ka)
)

two_way <- two_way |>
  mutate(
    pct_diff      = 100 * (est_age_ka - known_age_ka) / known_age_ka,
    in_interval66 = known_age_ka >= lo66 & known_age_ka <= hi66
  )

summary_vs_known <- two_way |>
  group_by(method) |>
  summarise(
    median_abs_pct_diff = median(abs(pct_diff)),
    rmse_log10_age       = sqrt(mean((log10(est_age_ka) - log10(known_age_ka))^2)),
    coverage66_pct        = 100 * mean(in_interval66),
    .groups = "drop"
  )

knitr::kable(summary_vs_known, digits = 2,
             col.names = c("Method", "Median abs. % diff vs. known age",
                           "RMSE (log10 age)", "66% interval coverage (%)"))
```

```{r plot_two_way, fig.height=4.5}
ggplot(two_way, aes(x = known_age_ka, y = est_age_ka, colour = method)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.6, size = 1.3) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~method) +
  labs(x = "Known (independently dated) age (ka)", y = "Estimated age (ka)",
       title = "Spreadsheet vs. Bayesian inversion, against known ages",
       subtitle = "Dashed line: 1:1; both panels share the same 128 samples",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "none")
```

```{r plot_two_way_diff, fig.height=4.5}
ggplot(two_way, aes(x = known_age_ka, y = pct_diff, colour = method)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.6, size = 1.3) +
  scale_x_log10() +
  facet_wrap(~method) +
  labs(x = "Known (independently dated) age (ka)",
       y = "Estimated minus known age, as % of known age",
       title = "Age differences from known ages, spreadsheet vs. Bayesian inversion",
       subtitle = "Dashed line: zero difference; same 128 samples as above",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "none")
```

```{r plot_two_way_density, fig.height=4}
ggplot(two_way, aes(x = pct_diff, fill = method, colour = method)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_density(alpha = 0.35, linewidth = 0.8) +
  labs(x = "Estimated minus known age, as % of known age", y = "Density",
       title = "Overall fit to known ages: distribution of % differences by method",
       fill = NULL, colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

Both methods track the known ages closely and, since they share the same
published calibration curves, are nearly indistinguishable by every metric
above: near-identical median absolute % difference, RMSE in log10(age),
and 66% interval coverage close to the nominal 66%. This is the expected
result of Stage 2's finding that GLS and the Bayesian inversion are
numerically almost equivalent once the calibration curve is fixed, and it
confirms the Bayesian inversion is a valid, independent check on the
spreadsheet calculator rather than a source of disagreement.

## Why the Bayesian inversion agrees with GLS

For a flat prior and Gaussian log-age residuals, the Bayesian posterior
mode and variance reduce algebraically to the GLS point estimate and
combined variance, given the same calibration curve. Because this notebook
supplies the Bayesian inversion with exactly the manuscript's own published
calibration curve (rather than re-estimating it), the two methods'
agreement here confirms the GLS formulation is a valid, deployable
equivalent of the Bayesian inversion, not a demonstration that Bayesian
calibration refitting would agree with GLS in general.

One simplification applies to the Bayesian inversion: the likelihood here
uses only the calibration-scale residual variance (`sigma_Asp`,
`sigma_Glu`), not the additional per-sample analytical-measurement-
uncertainty term the GLS calculator adds via each sample's replicate SE.
The manuscript finds that term contributes only a small amount to the
total variance on average, so this simplification has little practical
effect on the comparison above, though it can matter more for samples with
few replicates, especially young samples, where the GLS calculator's
explicit propagation of replicate uncertainty is an advantage over this
simplified Bayesian implementation.
