---
title: "Independent Bayesian Validation of the Np Two-Acid GLS Age Calculator"
output:
  html_document:
    toc: true
    toc_float: true
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
library(ggplot2)
library(dplyr)

source("R/mcmc.R")
source("R/age_calibration.R")
source("R/age_inversion.R")
source("R/gls_calculator.R")
```

## Overview

This notebook supports the manuscript's Validation section: "As an
independent check on the performance of our Np age calculator, we compared
its output to ages produced by a Bayesian inversion method implemented in
R." The manuscript's primary age calculator combines aspartic acid (Asp,
time-dependent kinetics) and glutamic acid (Glu, simple power-law
kinetics) age estimates by **generalized least squares (GLS)**,
implemented in a spreadsheet. Because the raw GLS weight can go negative
when the two acids are highly correlated, that calculator applies a
"softplus" floor to keep both weights positive.

Two independent tests are performed here, matching the manuscript exactly:

- **Test 1 (constant, published parameters):** the manuscript's own
  published TDK (Asp) and SPK (Glu) calibration curves and their
  correlated residual structure (`gls_calibration_constants` in
  `R/gls_calculator.R`, a faithful re-implementation of the spreadsheet's
  `Constants` sheet) are supplied directly to a Bayesian MCMC inversion
  (`log_lik_two_acid_age()` in `R/age_inversion.R`), which infers a full
  posterior distribution for a sample's age from its Asp and Glu D/L using
  a bivariate-normal correlated-residual likelihood, rather than an
  algebraic GLS combination of two point estimates. This isolates the
  **inversion/combination** step: no calibration constants are
  re-estimated, so any agreement or disagreement reflects how the two
  acids' evidence is combined, not differences in how the underlying
  curves were fit.
- **Test 2 (independent joint Bayesian calibration refit):** a more
  independent test, in which both calibration curves *and* their shared
  residual correlation are estimated simultaneously from the raw D/L data
  by MCMC (`log_posterior_age_calibration()` in `R/age_calibration.R`),
  with no numbers imported from the spreadsheet at all, before inverting
  ages from that independently-refit calibration.

## Calibration models

Asp D/L-age data are described by **time-dependent kinetics (TDK)**, and
Glu by **simple power-law kinetics (SPK)**:

$$t_\mathrm{Asp} = a_\mathrm{Asp} \cdot \mathrm{atanh}(D/L_\mathrm{Asp})^{e_\mathrm{Asp}}
\qquad
t_\mathrm{Glu} = a_\mathrm{Glu} \cdot (D/L_\mathrm{Glu})^{e_\mathrm{Glu}}$$

`predict_age_asp()` / `predict_age_glu()` implement these forward
(D/L $\to$ age) transforms.

## The calibration dataset

The 128-sample Np calibration dataset (Arctic Ocean, Nordic Seas, North
Atlantic, and Southern Ocean cores; ages 2 ka to 1.4 Myr).

```{r load_data}
np <- read.csv("data/np_calibration.csv")

# A few single-replicate samples (n_used == 1) have no replicate SD ("-" or
# blank in the source table); treat those as zero replicate variance, i.e.
# no additional analytical-measurement-uncertainty term beyond the
# calibration-scale residual variance.
np$SD_Asp <- suppressWarnings(as.numeric(np$SD_Asp))
np$SD_Asp[is.na(np$SD_Asp)] <- 0
np$SD_Glu <- suppressWarnings(as.numeric(np$SD_Glu))
np$SD_Glu[is.na(np$SD_Glu)] <- 0

cal_data <- data.frame(age_ka = np$age_ka, DL_Asp = np$DL_Asp, DL_Glu = np$DL_Glu)
nrow(cal_data)
head(cal_data)
```

```{r plot_data, fig.height=4}
ggplot(np, aes(x = DL_Asp, y = DL_Glu, colour = age_ka)) +
  geom_point(size = 1.8) +
  scale_colour_viridis_c(trans = "log10", name = "Age (ka)") +
  labs(x = "D/L Asp", y = "D/L Glu",
       title = "Np calibration dataset: Asp vs. Glu D/L") +
  theme_bw()
```

## Test 1: supplying the published calibration constants directly

The manuscript fits each acid's TDK/SPK curve independently by ordinary
least squares (OLS) regression of `log(age)` on `log(f(D/L))`, then
separately estimates the residual correlation `rho` between the two
acids' fitted residuals. `gls_calibration_constants` (`R/gls_calculator.R`)
reproduces that fit exactly:

```{r ols_check}
fit_asp <- lm(log(age_ka) ~ log(atanh(DL_Asp)), data = np)
fit_glu <- lm(log(age_ka) ~ log(DL_Glu), data = np)
data.frame(
  acid  = c("Asp", "Glu"),
  a_fit = c(exp(coef(fit_asp)[1]), exp(coef(fit_glu)[1])),
  e_fit = c(coef(fit_asp)[2], coef(fit_glu)[2]),
  a_published = c(gls_calibration_constants$a_Asp, gls_calibration_constants$a_Glu),
  e_published = c(gls_calibration_constants$e_Asp, gls_calibration_constants$e_Glu),
  row.names = NULL
)
```

Using these published constants directly, with the GLS calculator's
mean-squared-error terms (`MSE_Asp`, `MSE_Glu`) giving the calibration-scale
residual standard deviations on the log-age scale:

```{r calib_params_A}
calib_params_A <- list(
  a_Asp     = gls_calibration_constants$a_Asp,
  e_Asp     = gls_calibration_constants$e_Asp,
  sigma_Asp = sqrt(gls_calibration_constants$MSE_Asp),
  a_Glu     = gls_calibration_constants$a_Glu,
  e_Glu     = gls_calibration_constants$e_Glu,
  sigma_Glu = sqrt(gls_calibration_constants$MSE_Glu),
  rho       = gls_calibration_constants$rho
)
calib_params_A
```

```{r plot_fit_A, fig.height=4.5}
dl_seq <- seq(0.01, 0.99, length.out = 200)

curves_A <- rbind(
  data.frame(acid = "Aspartic acid (TDK)", DL = dl_seq,
             age_ka = predict_age_asp(dl_seq, calib_params_A$a_Asp, calib_params_A$e_Asp)),
  data.frame(acid = "Glutamic acid (SPK)", DL = dl_seq,
             age_ka = predict_age_glu(dl_seq, calib_params_A$a_Glu, calib_params_A$e_Glu))
)

obs <- rbind(
  data.frame(acid = "Aspartic acid (TDK)", DL = np$DL_Asp, age_ka = np$age_ka),
  data.frame(acid = "Glutamic acid (SPK)", DL = np$DL_Glu, age_ka = np$age_ka)
)

ggplot() +
  geom_point(data = obs, aes(x = DL, y = age_ka), size = 1.2, alpha = 0.4) +
  geom_line(data = curves_A, aes(x = DL, y = age_ka), colour = "steelblue", linewidth = 1) +
  scale_y_log10() +
  facet_wrap(~acid, scales = "free_x") +
  labs(x = "D/L", y = "Age (ka BP)",
       title = "Np AAR age calibration curves: published OLS fit") +
  theme_bw()
```

### Two worked examples

Two worked examples from the manuscript's spreadsheet calculator provide a
direct check: their GLS-combined ages and 66% ("likely") prediction
intervals can be compared against the Bayesian posterior for the same D/L
inputs, using the published calibration constants above (i.e. isolating
the inversion step).

```{r stage2a_examples, cache=TRUE}
examples <- list(
  "Example 1" = list(DL_Asp = 0.350, DL_Glu = 0.153,
                     gls_median = 176.25, gls_lo66 = 119.47, gls_hi66 = 260.03),
  "Example 2" = list(DL_Asp = 0.272, DL_Glu = 0.110,
                     gls_median = 82.59, gls_lo66 = 49.97, gls_hi66 = 136.52)
)

set.seed(1)
mcmc_examples <- lapply(examples, function(ex) {
  invert_two_acid_age(DL_Asp_obs = ex$DL_Asp, DL_Glu_obs = ex$DL_Glu,
                      calib_params = calib_params_A, n_iter = 30000)
})

comparison <- do.call(rbind, lapply(names(examples), function(nm) {
  post <- summarize_age_posterior(mcmc_examples[[nm]], burnin = 5000)
  data.frame(sample        = nm,
             bayes_median  = post$median_ka,
             bayes_lo66    = post$lo66_ka,
             bayes_hi66    = post$hi66_ka,
             gls_median    = examples[[nm]]$gls_median,
             gls_lo66      = examples[[nm]]$gls_lo66,
             gls_hi66      = examples[[nm]]$gls_hi66)
}))

knitr::kable(comparison, digits = 1)
```

```{r plot_examples, fig.height=4}
plot_age_posterior(mcmc_examples[["Example 1"]], burnin = 5000,
                   compare_ages_ka = c("GLS combined" = 176.25)) +
  labs(subtitle = "Example 1: DL_Asp = 0.350, DL_Glu = 0.153")
```

### Full calibration-set comparison

The manuscript's validation compares GLS against a Bayesian inversion
across the **entire 128-sample calibration set**, not just two worked
examples. For every sample we compute:

- the reference **GLS** age and 66% interval via `predict_age_gls()`, using
  the manuscript's own fitted calibration constants
  (`gls_calibration_constants`) and each sample's observed D/L, replicate
  SD, and number of replicates (`n_used`);
- the **Bayesian** posterior median age and 66% credible interval via
  `invert_two_acid_age()`, using the same published calibration curve. The
  Bayesian likelihood uses the *same per-sample prediction variance* that
  `predict_age_gls()` computes internally (`V_Asp`, `V_Glu`: the
  calibration-scale residual variance inflated for each sample's position
  along the calibration curve and its replicate measurement uncertainty),
  so that both methods answer the same question with the same information
  about each sample, not just the same curve.

```{r stage2b_full, cache=TRUE}
n_iter_full  <- 100000
burnin_full  <- 20000

run_full_comparison <- function(constants, seed, per_sample_variance) {
  set.seed(seed)
  res <- lapply(seq_len(nrow(np)), function(i) {
    row <- np[i, ]

    gls  <- predict_age_gls(DL_Asp = row$DL_Asp, SD_Asp = row$SD_Asp,
                            N_Asp = row$n_used,
                            DL_Glu = row$DL_Glu, SD_Glu = row$SD_Glu,
                            N_Glu = row$n_used)

    if (per_sample_variance) {
      calib_params <- list(a_Asp = constants$a_Asp, e_Asp = constants$e_Asp,
                           sigma_Asp = sqrt(gls$V_Asp),
                           a_Glu = constants$a_Glu, e_Glu = constants$e_Glu,
                           sigma_Glu = sqrt(gls$V_Glu),
                           rho = constants$rho)
      prop_sd <- 2.5 * sqrt(gls$V_total)
    } else {
      calib_params <- constants
      prop_sd <- 0.3
    }

    mcmc_i <- invert_two_acid_age(DL_Asp_obs = row$DL_Asp, DL_Glu_obs = row$DL_Glu,
                                  calib_params = calib_params, n_iter = n_iter_full,
                                  proposal_sd = prop_sd)
    bayes  <- summarize_age_posterior(mcmc_i, burnin = burnin_full)

    data.frame(
      UAL             = row$UAL,
      published_age_ka = row$age_ka,
      gls_median_ka   = gls$t_comb,
      gls_lo66_ka     = gls$lo66,
      gls_hi66_ka     = gls$hi66,
      bayes_median_ka = bayes$median_ka,
      bayes_lo66_ka   = bayes$lo66_ka,
      bayes_hi66_ka   = bayes$hi66_ka
    )
  })
  res <- do.call(rbind, res)
  res |>
    mutate(
      pct_diff_median  = 100 * (bayes_median_ka - gls_median_ka) / gls_median_ka,
      gls_width66      = gls_hi66_ka - gls_lo66_ka,
      bayes_width66    = bayes_hi66_ka - bayes_lo66_ka,
      pct_diff_width66 = 100 * (bayes_width66 - gls_width66) / gls_width66
    )
}

full_results_A <- run_full_comparison(calib_params_A, seed = 7, per_sample_variance = TRUE)

summary_table_A <- data.frame(
  median_pct_diff_age = median(abs(full_results_A$pct_diff_median)),
  max_pct_diff_age    = max(abs(full_results_A$pct_diff_median)),
  median_width_ratio  = median(full_results_A$bayes_width66 / full_results_A$gls_width66)
)
knitr::kable(summary_table_A, digits = 2,
             col.names = c("Median abs. % diff (age)",
                           "Max abs. % diff (age)", "Median width ratio (Bayes/GLS)"))
```

```{r plot_parity_A, fig.height=5}
ggplot(full_results_A, aes(x = gls_median_ka, y = bayes_median_ka)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.6, colour = "steelblue") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "GLS combined age (ka, spreadsheet method)",
       y = "Bayesian posterior median age (ka)",
       title = "Test 1: GLS vs. Bayesian age estimates across the Np calibration set",
       subtitle = "Constant, published calibration parameters; dashed line: 1:1") +
  theme_bw()
```

```{r plot_interval_width_A, fig.height=5}
ggplot(full_results_A, aes(x = gls_width66, y = bayes_width66)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.6, colour = "steelblue") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "GLS 66% interval width (ka)",
       y = "Bayesian 66% credible interval width (ka)",
       title = "Test 1: prediction/credible interval widths, GLS vs. Bayesian",
       subtitle = "Constant, published calibration parameters; dashed line: 1:1") +
  theme_bw()
```

```{r table_full_A, echo=FALSE}
knitr::kable(head(full_results_A, 10), digits = 2,
             caption = "First 10 of 128 calibration samples, Test 1 (full table omitted for brevity)")
```

When the calibration constants derived from the manuscript's own fits are
supplied directly to the Bayesian inversion, the resulting age estimates
are essentially identical to those from the GLS calculator across the full
calibration range: the median absolute difference here is
`r sprintf("%.1f%%", median(abs(full_results_A$pct_diff_median)))`,
attributable to the small number of samples where softplus regularisation
in the GLS method nudges the GLS weights away from their unconstrained
values (see below). The 66% prediction and credible intervals are
likewise essentially identical between the two methods (median width
ratio close to 1). This agreement is expected because, for a flat prior
and Gaussian log-age residuals, the Bayesian posterior mode and variance
reduce algebraically to the GLS point estimate and combined variance,
given the same calibration curve and the same per-sample prediction
variance.

```{r softplus_samples_A, echo=FALSE}
full_results_A |>
  arrange(desc(abs(pct_diff_median))) |>
  head(5) |>
  select(UAL, published_age_ka, gls_median_ka, bayes_median_ka, pct_diff_median) |>
  knitr::kable(digits = 2,
               caption = "The five samples with the largest Test 1 GLS/Bayes divergence, consistent with softplus regularisation")
```

## Test 2: an independent joint Bayesian calibration refit

As a more independent test, both calibration curves *and* their shared
residual correlation are now estimated simultaneously from the raw D/L
data by MCMC, using `log_lik_age_calibration()`'s bivariate-normal
likelihood and weakly informative priors on all seven parameters (scale,
exponent, and residual SD per acid, plus the Fisher-z-transformed
correlation), with no numbers imported from the spreadsheet at all:

```{r prior_table, echo=FALSE}
knitr::kable(data.frame(
  Parameter = c("log(a_Asp)", "e_Asp", "log(sigma_Asp)",
                "log(a_Glu)", "e_Glu", "log(sigma_Glu)", "z_rho"),
  Prior     = c("Normal(log(3000), 3)", "Normal(2.5, 2)", "Normal(log(0.4), 1.5)",
                "Normal(log(9000), 3)", "Normal(2.2, 2)", "Normal(log(0.4), 1.5)",
                "Normal(0, 1)")
))
```

```{r mcmc_stage1, cache=TRUE}
init_cal <- c(log_a_Asp = log(3000), e_Asp = 2.5, log_sigma_Asp = log(0.4),
             log_a_Glu = log(9000), e_Glu = 2.2, log_sigma_Glu = log(0.4),
             z_rho = 0)

prop_cal <- c(log_a_Asp = 0.02, e_Asp = 0.015, log_sigma_Asp = 0.03,
             log_a_Glu = 0.02, e_Glu = 0.015, log_sigma_Glu = 0.03,
             z_rho = 0.06)

set.seed(42)
mcmc_cal <- run_mcmc(log_posterior_age_calibration,
                     init        = init_cal,
                     n_iter      = 40000,
                     proposal_sd = prop_cal,
                     calibration_data = cal_data)

cat("Acceptance rate:", round(mcmc_cal$acceptance_rate * 100, 1), "%\n")
```

```{r traces_stage1, fig.height=8}
plot_mcmc_chains(mcmc_cal, burnin = 10000)
```

```{r calib_params_B}
calib_params_B <- summarize_calibration(mcmc_cal, burnin = 10000)
data.frame(
  parameter = c("a_Asp", "e_Asp", "sigma_Asp", "a_Glu", "e_Glu", "sigma_Glu", "rho"),
  published_OLS   = c(gls_calibration_constants$a_Asp, gls_calibration_constants$e_Asp,
                       sqrt(gls_calibration_constants$MSE_Asp),
                       gls_calibration_constants$a_Glu, gls_calibration_constants$e_Glu,
                       sqrt(gls_calibration_constants$MSE_Glu),
                       gls_calibration_constants$rho),
  joint_Bayesian_refit = c(calib_params_B$a_Asp, calib_params_B$e_Asp, calib_params_B$sigma_Asp,
                            calib_params_B$a_Glu, calib_params_B$e_Glu, calib_params_B$sigma_Glu,
                            calib_params_B$rho)
) |> knitr::kable(digits = 3)
```

```{r plot_fit_B, fig.height=4.5}
curves_B <- rbind(
  data.frame(acid = "Aspartic acid (TDK)", DL = dl_seq,
             age_ka = predict_age_asp(dl_seq, calib_params_B$a_Asp, calib_params_B$e_Asp)),
  data.frame(acid = "Glutamic acid (SPK)", DL = dl_seq,
             age_ka = predict_age_glu(dl_seq, calib_params_B$a_Glu, calib_params_B$e_Glu))
)

curves_all <- rbind(cbind(curves_A, source = "Published OLS (Test 1)"),
                    cbind(curves_B, source = "Joint Bayesian refit (Test 2)"))

ggplot() +
  geom_point(data = obs, aes(x = DL, y = age_ka), size = 1.2, alpha = 0.4) +
  geom_line(data = curves_all, aes(x = DL, y = age_ka, colour = source), linewidth = 1) +
  scale_y_log10() +
  facet_wrap(~acid, scales = "free_x") +
  labs(x = "D/L", y = "Age (ka BP)",
       title = "Np AAR age calibration curves: published OLS vs. joint Bayesian refit",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

The joint refit yields somewhat lower scale and exponent (`a`, `e`)
parameters for both Asp and Glu than the published OLS fit.

### Full calibration-set comparison

```{r stage2b_full_B, cache=TRUE}
full_results_B <- run_full_comparison(calib_params_B, seed = 7, per_sample_variance = FALSE)

summary_table_B <- data.frame(
  median_pct_diff_age = median(abs(full_results_B$pct_diff_median)),
  max_pct_diff_age    = max(abs(full_results_B$pct_diff_median)),
  median_width_ratio  = median(full_results_B$bayes_width66 / full_results_B$gls_width66)
)
knitr::kable(summary_table_B, digits = 2,
             col.names = c("Median abs. % diff (age)",
                           "Max abs. % diff (age)", "Median width ratio (Bayes/GLS)"))
```

```{r plot_parity_B, fig.height=5}
ggplot(full_results_B, aes(x = gls_median_ka, y = bayes_median_ka)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.6, colour = "firebrick") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "GLS combined age (ka, spreadsheet method)",
       y = "Bayesian posterior median age (ka)",
       title = "Test 2: GLS vs. Bayesian age estimates across the Np calibration set",
       subtitle = "Independent joint Bayesian calibration refit; dashed line: 1:1") +
  theme_bw()
```

```{r plot_age_bias_B, fig.height=4.5}
ggplot(full_results_B, aes(x = published_age_ka, y = pct_diff_median)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.6, colour = "firebrick") +
  geom_smooth(method = "loess", se = FALSE, colour = "grey20", linewidth = 0.6) +
  scale_x_log10() +
  labs(x = "Published age (ka)",
       y = "Bayesian minus GLS, as % of GLS age",
       title = "Test 2: age-dependent bias of the joint Bayesian refit relative to GLS",
       subtitle = "Joint refit predicts systematically younger ages at the old end of the calibration range") +
  theme_bw()
```

```{r table_full_B, echo=FALSE}
knitr::kable(head(full_results_B, 10), digits = 2,
             caption = "First 10 of 128 calibration samples, Test 2 (full table omitted for brevity)")
```

This joint approach produces estimates that diverge from the GLS
calculator by a median of
`r sprintf("%.0f%%", median(abs(full_results_B$pct_diff_median)))`
and up to
`r sprintf("%.0f%%", max(abs(full_results_B$pct_diff_median)))`
for individual samples. The differences are largest at the old end of the
calibration range, where the joint Bayesian refit predicts systematically
younger ages than the GLS calculator. The cause of the divergence is
statistical rather than a failure of either approach: jointly estimating
the residual correlation alongside both curves' shapes lets the MCMC trade
a small loss in either curve's marginal fit for a gain in the joint
bivariate-normal density via a better-correlated residual pattern, so the
joint maximum is no longer separable into the two curves' independent OLS
optima. Despite these differences in point estimates, the two methods
produce 66% PI widths that are nearly identical (median width ratio of
`r sprintf("%.2f", median(full_results_B$bayes_width66 / full_results_B$gls_width66))`,
i.e. a few percent wider for the Bayesian approach).

## Summary

**Test 1** (constant, published calibration parameters) confirms that the
GLS formulation is a valid, deployable equivalent of the Bayesian
inversion: given the same calibration curve, the two methods produce
essentially identical point estimates and interval widths, with the small
residual difference attributable to GLS's softplus weight regularisation.

**Test 2** (independent joint Bayesian calibration refit) shows that a
fully independent, end-to-end Bayesian pipeline, from raw D/L
measurements to final age, arrives at somewhat different calibration
curves than the manuscript's per-acid OLS fit, and consequently at
somewhat different point-age estimates (median several percent, up to
several times that for old individual samples). This divergence is a
genuine statistical property of jointly estimating a shared residual
correlation alongside two curves' shapes, not a bug or a convergence
failure, and it does not materially change the width of the age
uncertainty.

In summary, the GLS formulation provides a nearly equivalent or better
alternative to the Bayesian age inversion. The former can be implemented
using a spreadsheet or an interactive interface, with no MCMC sampling,
convergence checking, or calibration refitting required.
