---
title: "Independent Bayesian Validation of the Np Two-Acid GLS Age Calculator"
output:
  html_document:
    toc: true
    toc_float: true
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 5
)
library(ggplot2)
library(dplyr)

source("R/mcmc.R")
source("R/age_calibration.R")
source("R/age_inversion.R")
source("R/gls_calculator.R")
```

## Overview

This notebook supports section 4.1 ("Validation by independent Bayesian
method") of the Np AAR age-calculator manuscript. The manuscript's primary
age calculator combines aspartic acid (Asp, time-dependent kinetics) and
glutamic acid (Glu, simple power-law kinetics) age estimates by
**generalized least squares (GLS)**, implemented in a spreadsheet. Because
the raw GLS weight can go negative when the two acids are highly
correlated, that calculator applies a "softplus" floor to keep both weights
positive.

As an independent check, this notebook performs a **Bayesian MCMC
inversion** of the same two-acid evidence: given the manuscript's own
fitted TDK (Asp) and SPK (Glu) calibration curves and their correlated
residual structure (`R/gls_calculator.R`, a faithful re-implementation of
the spreadsheet's `Constants` sheet), it infers a full posterior
distribution for a sample's age from its Asp and Glu D/L using a bivariate
normal likelihood (`log_lik_two_acid_age()` in `R/age_inversion.R`), rather
than an algebraic GLS combination of two point estimates. Because this is
an ordinary Bayesian posterior, the two acids combine coherently for any
value of the residual correlation with no floor or smoothing constant
needed.

This validation isolates the **inversion/combination** step specifically:
the main comparison (Stage 2) starts from the same fitted calibration
curves as the GLS calculator, so any agreement (or disagreement) there
reflects how the two acids' evidence is combined, not differences in how
the underlying curves were fit. Stage 1 below additionally investigates a
**second way** to obtain those calibration curves — an independent,
from-scratch joint Bayesian MCMC refit — because its posterior constants
turn out to diverge non-trivially from the manuscript's OLS fit. That
divergence, why it happens, and what it does to the downstream age
estimates and uncertainties, is examined explicitly rather than set aside.

## Calibration models

Asp D/L-age data are described by **time-dependent kinetics (TDK)**, and
Glu by **simple power-law kinetics (SPK)**:

$$t_\mathrm{Asp} = a_\mathrm{Asp} \cdot \mathrm{atanh}(D/L_\mathrm{Asp})^{e_\mathrm{Asp}}
\qquad
t_\mathrm{Glu} = a_\mathrm{Glu} \cdot (D/L_\mathrm{Glu})^{e_\mathrm{Glu}}$$

`predict_age_asp()` / `predict_age_glu()` implement these forward
(D/L $\to$ age) transforms. The scale/exponent constants and residual
correlation (`gls_calibration_constants` in `R/gls_calculator.R`) are taken
directly from the manuscript's own OLS fit of the 128-sample calibration
set, not re-estimated here.

## The calibration dataset

The 128-sample Np calibration dataset (Arctic Ocean, Nordic Seas, North
Atlantic, and Southern Ocean cores; ages 2 ka to 1.4 Myr).

```{r load_data}
np <- read.csv("data/np_calibration.csv")

# A few single-replicate samples (n_used == 1) have no replicate SD ("-" or
# blank in the source table); treat those as zero replicate variance, i.e.
# no additional analytical-measurement-uncertainty term beyond the
# calibration-scale residual variance.
np$SD_Asp <- suppressWarnings(as.numeric(np$SD_Asp))
np$SD_Asp[is.na(np$SD_Asp)] <- 0
np$SD_Glu <- suppressWarnings(as.numeric(np$SD_Glu))
np$SD_Glu[is.na(np$SD_Glu)] <- 0

cal_data <- data.frame(age_ka = np$age_ka, DL_Asp = np$DL_Asp, DL_Glu = np$DL_Glu)
nrow(cal_data)
head(cal_data)
```

```{r plot_data, fig.height=4}
ggplot(np, aes(x = DL_Asp, y = DL_Glu, colour = age_ka)) +
  geom_point(size = 1.8) +
  scale_colour_viridis_c(trans = "log10", name = "Age (ka)") +
  labs(x = "D/L Asp", y = "D/L Glu",
       title = "Np calibration dataset: Asp vs. Glu D/L") +
  theme_bw()
```

## Stage 1: two ways to obtain the calibration constants

### Approach A: the manuscript's published OLS constants

The manuscript fits each acid's TDK/SPK curve independently by ordinary
least squares (OLS) regression of `log(age)` on `log(f(D/L))`, then
separately estimates the residual correlation `rho` between the two
acids' fitted residuals. `gls_calibration_constants` (`R/gls_calculator.R`)
reproduces that fit exactly:

```{r ols_check}
fit_asp <- lm(log(age_ka) ~ log(atanh(DL_Asp)), data = np)
fit_glu <- lm(log(age_ka) ~ log(DL_Glu), data = np)
data.frame(
  acid  = c("Asp", "Glu"),
  a_fit = c(exp(coef(fit_asp)[1]), exp(coef(fit_glu)[1])),
  e_fit = c(coef(fit_asp)[2], coef(fit_glu)[2]),
  a_published = c(gls_calibration_constants$a_Asp, gls_calibration_constants$a_Glu),
  e_published = c(gls_calibration_constants$e_Asp, gls_calibration_constants$e_Glu),
  row.names = NULL
)
```

Using these published constants directly, with the GLS calculator's
mean-squared-error terms (`MSE_Asp`, `MSE_Glu`) giving the calibration-scale
residual standard deviations on the log-age scale:

```{r calib_params_A}
calib_params_A <- list(
  a_Asp     = gls_calibration_constants$a_Asp,
  e_Asp     = gls_calibration_constants$e_Asp,
  sigma_Asp = sqrt(gls_calibration_constants$MSE_Asp),
  a_Glu     = gls_calibration_constants$a_Glu,
  e_Glu     = gls_calibration_constants$e_Glu,
  sigma_Glu = sqrt(gls_calibration_constants$MSE_Glu),
  rho       = gls_calibration_constants$rho
)
calib_params_A
```

### Approach B: an independent, from-scratch joint Bayesian refit

Instead of taking the two acids' curves as given, this approach fits both
curves *and* their shared residual correlation jointly in a single MCMC,
using `log_lik_age_calibration()`'s bivariate-normal likelihood and weakly
informative priors on all seven parameters (scale, exponent, and residual
SD per acid, plus the Fisher-z-transformed correlation):

```{r prior_table, echo=FALSE}
knitr::kable(data.frame(
  Parameter = c("log(a_Asp)", "e_Asp", "log(sigma_Asp)",
                "log(a_Glu)", "e_Glu", "log(sigma_Glu)", "z_rho"),
  Prior     = c("Normal(log(3000), 3)", "Normal(2.5, 2)", "Normal(log(0.4), 1.5)",
                "Normal(log(9000), 3)", "Normal(2.2, 2)", "Normal(log(0.4), 1.5)",
                "Normal(0, 1)")
))
```

```{r mcmc_stage1, cache=TRUE}
init_cal <- c(log_a_Asp = log(3000), e_Asp = 2.5, log_sigma_Asp = log(0.4),
             log_a_Glu = log(9000), e_Glu = 2.2, log_sigma_Glu = log(0.4),
             z_rho = 0)

prop_cal <- c(log_a_Asp = 0.02, e_Asp = 0.015, log_sigma_Asp = 0.03,
             log_a_Glu = 0.02, e_Glu = 0.015, log_sigma_Glu = 0.03,
             z_rho = 0.06)

set.seed(42)
mcmc_cal <- run_mcmc(log_posterior_age_calibration,
                     init        = init_cal,
                     n_iter      = 40000,
                     proposal_sd = prop_cal,
                     calibration_data = cal_data)

cat("Acceptance rate:", round(mcmc_cal$acceptance_rate * 100, 1), "%\n")
```

```{r traces_stage1, fig.height=8}
plot_mcmc_chains(mcmc_cal, burnin = 10000)
```

```{r calib_params_B}
calib_params_B <- summarize_calibration(mcmc_cal, burnin = 10000)
calib_params_B
```

### Why do Approach A and Approach B differ?

Approach B's constants diverge from Approach A's by more than MCMC noise:
repeating the fit from different random seeds converges to the same
values (acceptance rate ~27%, stable across seeds), so this is not a
convergence problem. The divergence is also not simply "Bayesian vs.
frequentist," since a **per-acid** OLS refit (above) reproduces the
published constants almost exactly. The difference is specifically caused
by estimating the **residual correlation jointly with both curves' shapes**:
refitting with `z_rho` fixed at exactly zero (i.e. the same joint MCMC, but
with the two acids' likelihoods forced to be independent) recovers
constants close to the published OLS values again:

```{r rho0_diagnostic, cache=TRUE}
log_post_rho0 <- function(params, calibration_data) {
  log_posterior_age_calibration(c(params, z_rho = 0), calibration_data)
}

init_rho0 <- init_cal[names(init_cal) != "z_rho"]
prop_rho0 <- prop_cal[names(prop_cal) != "z_rho"]

set.seed(1)
mcmc_rho0 <- run_mcmc(log_post_rho0, init = init_rho0, n_iter = 40000,
                      proposal_sd = prop_rho0, calibration_data = cal_data)
post_rho0 <- colMeans(utils::tail(mcmc_rho0$samples, nrow(mcmc_rho0$samples) - 10000))

data.frame(
  source = c("Published OLS", "Joint MCMC (rho free)", "Joint MCMC (rho = 0)"),
  a_Asp  = c(gls_calibration_constants$a_Asp, calib_params_B$a_Asp, exp(post_rho0[["log_a_Asp"]])),
  e_Asp  = c(gls_calibration_constants$e_Asp, calib_params_B$e_Asp, post_rho0[["e_Asp"]]),
  a_Glu  = c(gls_calibration_constants$a_Glu, calib_params_B$a_Glu, exp(post_rho0[["log_a_Glu"]])),
  e_Glu  = c(gls_calibration_constants$e_Glu, calib_params_B$e_Glu, post_rho0[["e_Glu"]])
)
```

In other words, once `rho` is free to be jointly estimated, the MCMC can
trade a small loss in either curve's marginal fit for a gain in the joint
bivariate-normal density via a better-correlated residual pattern — the
joint maximum is no longer separable into the two curves' independent OLS
optima. This is a genuine statistical property of the joint-likelihood
approach, not a bug or a convergence failure.

### Fitted curves vs. data (both approaches)

```{r plot_fit, fig.height=4.5}
dl_seq <- seq(0.01, 0.99, length.out = 200)

curve_set <- function(cp, label) {
  curves <- rbind(
    data.frame(acid = "Aspartic acid (TDK)", DL = dl_seq,
               age_ka = predict_age_asp(dl_seq, cp$a_Asp, cp$e_Asp)),
    data.frame(acid = "Glutamic acid (SPK)", DL = dl_seq,
               age_ka = predict_age_glu(dl_seq, cp$a_Glu, cp$e_Glu))
  )
  curves$approach <- label
  curves
}
curves <- rbind(curve_set(calib_params_A, "A: published OLS"),
                curve_set(calib_params_B, "B: joint Bayesian refit"))

obs <- rbind(
  data.frame(acid = "Aspartic acid (TDK)", DL = np$DL_Asp, age_ka = np$age_ka),
  data.frame(acid = "Glutamic acid (SPK)", DL = np$DL_Glu, age_ka = np$age_ka)
)

ggplot() +
  geom_point(data = obs, aes(x = DL, y = age_ka), size = 1.2, alpha = 0.4) +
  geom_line(data = curves, aes(x = DL, y = age_ka, colour = approach), linewidth = 1) +
  scale_y_log10() +
  facet_wrap(~acid, scales = "free_x") +
  labs(x = "D/L", y = "Age (ka BP)",
       title = "Np AAR age calibration curves: published OLS vs. joint Bayesian refit",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

## Stage 2a: two worked examples

Two worked examples from the manuscript's spreadsheet calculator provide a
direct check: their GLS-combined ages and 66% ("likely") prediction
intervals can be compared against the Bayesian posterior for the same D/L
inputs, using **Approach A**'s calibration constants (i.e. isolating the
inversion step, as in the main validation).

```{r stage2a_examples, cache=TRUE}
examples <- list(
  "Example 1" = list(DL_Asp = 0.350, DL_Glu = 0.153,
                     gls_median = 176.25, gls_lo66 = 119.47, gls_hi66 = 260.03),
  "Example 2" = list(DL_Asp = 0.272, DL_Glu = 0.110,
                     gls_median = 82.59, gls_lo66 = 49.97, gls_hi66 = 136.52)
)

set.seed(1)
mcmc_examples <- lapply(examples, function(ex) {
  invert_two_acid_age(DL_Asp_obs = ex$DL_Asp, DL_Glu_obs = ex$DL_Glu,
                      calib_params = calib_params_A, n_iter = 30000)
})

comparison <- do.call(rbind, lapply(names(examples), function(nm) {
  post <- summarize_age_posterior(mcmc_examples[[nm]], burnin = 5000)
  data.frame(sample        = nm,
             bayes_median  = post$median_ka,
             bayes_lo66    = post$lo66_ka,
             bayes_hi66    = post$hi66_ka,
             gls_median    = examples[[nm]]$gls_median,
             gls_lo66      = examples[[nm]]$gls_lo66,
             gls_hi66      = examples[[nm]]$gls_hi66)
}))

knitr::kable(comparison, digits = 1)
```

```{r plot_examples, fig.height=4}
plot_age_posterior(mcmc_examples[["Example 1"]], burnin = 5000,
                   compare_ages_ka = c("GLS combined" = 176.25)) +
  labs(subtitle = "Example 1: DL_Asp = 0.350, DL_Glu = 0.153")
```

## Stage 2b: full calibration-set comparison

The real validation in the manuscript compares GLS against a Bayesian
inversion across the **entire 128-sample calibration set**, not just two
worked examples. This is repeated for **both** calibration approaches from
Stage 1, so the effect of the calibration-constant choice on the
GLS-vs.-Bayesian agreement is visible directly. For every sample we compute:

- the reference **GLS** age and 66% interval via `predict_age_gls()`, using
  the manuscript's own fitted calibration constants
  (`gls_calibration_constants`) and each sample's observed D/L, replicate
  SD, and number of replicates (`n_used`) — this is fixed across both runs;
- the **Bayesian** posterior median age and 66% credible interval via
  `invert_two_acid_age()`, using either `calib_params_A` (published OLS) or
  `calib_params_B` (joint Bayesian refit).

```{r stage2b_full, cache=TRUE}
n_iter_full  <- 6000
burnin_full  <- 1500

run_full_comparison <- function(calib_params, seed) {
  set.seed(seed)
  res <- lapply(seq_len(nrow(np)), function(i) {
    row <- np[i, ]

    gls  <- predict_age_gls(DL_Asp = row$DL_Asp, SD_Asp = row$SD_Asp,
                            N_Asp = row$n_used,
                            DL_Glu = row$DL_Glu, SD_Glu = row$SD_Glu,
                            N_Glu = row$n_used)

    mcmc_i <- invert_two_acid_age(DL_Asp_obs = row$DL_Asp, DL_Glu_obs = row$DL_Glu,
                                  calib_params = calib_params, n_iter = n_iter_full)
    bayes  <- summarize_age_posterior(mcmc_i, burnin = burnin_full)

    data.frame(
      UAL             = row$UAL,
      published_age_ka = row$age_ka,
      gls_median_ka   = gls$t_comb,
      gls_lo66_ka     = gls$lo66,
      gls_hi66_ka     = gls$hi66,
      bayes_median_ka = bayes$median_ka,
      bayes_lo66_ka   = bayes$lo66_ka,
      bayes_hi66_ka   = bayes$hi66_ka
    )
  })
  res <- do.call(rbind, res)
  res |>
    mutate(
      pct_diff_median  = 100 * abs(bayes_median_ka - gls_median_ka) / gls_median_ka,
      gls_width66      = gls_hi66_ka - gls_lo66_ka,
      bayes_width66    = bayes_hi66_ka - bayes_lo66_ka,
      pct_diff_width66 = 100 * abs(bayes_width66 - gls_width66) / gls_width66
    )
}

full_results_A <- run_full_comparison(calib_params_A, seed = 7)
full_results_B <- run_full_comparison(calib_params_B, seed = 7)

summary_table <- data.frame(
  approach = c("A: published OLS", "B: joint Bayesian refit"),
  median_pct_diff_age    = c(median(full_results_A$pct_diff_median),
                             median(full_results_B$pct_diff_median)),
  max_pct_diff_age       = c(max(full_results_A$pct_diff_median),
                             max(full_results_B$pct_diff_median)),
  median_width_ratio     = c(median(full_results_A$bayes_width66 / full_results_A$gls_width66),
                             median(full_results_B$bayes_width66 / full_results_B$gls_width66))
)
knitr::kable(summary_table, digits = 2,
             col.names = c("Calibration approach", "Median abs. % diff (age)",
                           "Max abs. % diff (age)", "Median width ratio (Bayes/GLS)"))
```

```{r plot_parity, fig.height=5}
full_results <- rbind(
  cbind(full_results_A, approach = "A: published OLS"),
  cbind(full_results_B, approach = "B: joint Bayesian refit")
)

ggplot(full_results, aes(x = gls_median_ka, y = bayes_median_ka, colour = approach)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.6) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "GLS combined age (ka, spreadsheet method)",
       y = "Bayesian posterior median age (ka)",
       title = "GLS vs. Bayesian age estimates across the Np calibration set",
       subtitle = "Dashed line: 1:1", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r plot_interval_width, fig.height=5}
ggplot(full_results, aes(x = gls_width66, y = bayes_width66, colour = approach)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.6) +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "GLS 66% interval width (ka)",
       y = "Bayesian 66% credible interval width (ka)",
       title = "Prediction/credible interval widths: GLS vs. Bayesian",
       subtitle = "Dashed line: 1:1", colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r table_full, echo=FALSE}
knitr::kable(head(full_results_A, 10), digits = 2,
             caption = "First 10 of 128 calibration samples, Approach A (full table omitted for brevity)")
```

**Point estimates:** Approach A (published OLS constants) agrees with GLS
closely — a median absolute difference of a few percent, driven almost
entirely by the handful of samples where the GLS softplus regularisation
nudges the raw (possibly negative) GLS weight away from its unconstrained
value, a step the Bayesian posterior does not need. Approach B (joint
Bayesian refit) agrees far less well on point estimates — its median
absolute difference is several times larger — because its underlying curve
constants (particularly `a_Glu`/`e_Glu`) are themselves shifted from the
GLS calculator's curves, for the reason identified in Stage 1 above. This
is a difference in *calibration curve*, not in how the two acids' evidence
is combined once a curve is fixed.

**Uncertainty (interval widths):** despite the point-estimate gap, Approach
B's 66% credible interval widths track the GLS 66% interval widths about as
closely as Approach A's do (both ratios near 1). This is because Approach
B's posterior residual SDs (`sigma_Asp`, `sigma_Glu`) — which set the
*width* of the age uncertainty — turn out to be similar in magnitude to
Approach A's, even though the *location* of the curves (`a`, `e`) differs
enough to shift the point estimate. In other words, refitting the
calibration curves jointly mainly relocates the age estimate; it does not
by itself make the age noticeably more or less certain.

## Stage 2c: all three approaches vs. independently known ages

The comparisons above (Stage 2b) check the methods against each other. A
more direct question is how each one performs against the calibration
set's own independently established ages (`age_ka`, from radiocarbon,
oxygen-isotope stratigraphy, etc., depending on the source study), i.e.
the "known ages" that the calibration is built from. Three approaches are
compared for every one of the 128 samples:

- **Spreadsheet (GLS)**: the manuscript's published calculator
  (`predict_age_gls()`), combining the two acids by generalized least
  squares with the softplus floor.
- **Bayesian, constant parameters**: the MCMC age inversion using
  Approach A's calibration constants (the manuscript's own published OLS
  fit, taken as fixed).
- **Bayesian, estimated parameters**: the MCMC age inversion using
  Approach B's calibration constants (the independent joint Bayesian
  refit from Stage 1).

```{r stage3_known_ages}
three_way <- rbind(
  data.frame(UAL = full_results_A$UAL,
             known_age_ka = full_results_A$published_age_ka,
             method   = "Spreadsheet (GLS)",
             est_age_ka = full_results_A$gls_median_ka,
             lo66 = full_results_A$gls_lo66_ka,
             hi66 = full_results_A$gls_hi66_ka),
  data.frame(UAL = full_results_A$UAL,
             known_age_ka = full_results_A$published_age_ka,
             method   = "Bayesian, constant parameters (A)",
             est_age_ka = full_results_A$bayes_median_ka,
             lo66 = full_results_A$bayes_lo66_ka,
             hi66 = full_results_A$bayes_hi66_ka),
  data.frame(UAL = full_results_B$UAL,
             known_age_ka = full_results_B$published_age_ka,
             method   = "Bayesian, estimated parameters (B)",
             est_age_ka = full_results_B$bayes_median_ka,
             lo66 = full_results_B$bayes_lo66_ka,
             hi66 = full_results_B$bayes_hi66_ka)
)

three_way <- three_way |>
  mutate(
    pct_diff      = 100 * (est_age_ka - known_age_ka) / known_age_ka,
    in_interval66 = known_age_ka >= lo66 & known_age_ka <= hi66
  )

summary_vs_known <- three_way |>
  group_by(method) |>
  summarise(
    median_abs_pct_diff = median(abs(pct_diff)),
    rmse_log10_age       = sqrt(mean((log10(est_age_ka) - log10(known_age_ka))^2)),
    coverage66_pct        = 100 * mean(in_interval66),
    .groups = "drop"
  )

knitr::kable(summary_vs_known, digits = 2,
             col.names = c("Method", "Median abs. % diff vs. known age",
                           "RMSE (log10 age)", "66% interval coverage (%)"))
```

```{r plot_three_way, fig.height=4.5}
ggplot(three_way, aes(x = known_age_ka, y = est_age_ka, colour = method)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.6, size = 1.3) +
  scale_x_log10() +
  scale_y_log10() +
  facet_wrap(~method) +
  labs(x = "Known (independently dated) age (ka)", y = "Estimated age (ka)",
       title = "Spreadsheet vs. both Bayesian approaches, against known ages",
       subtitle = "Dashed line: 1:1; all three panels share the same 128 samples",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "none")
```

```{r plot_three_way_diff, fig.height=4.5}
ggplot(three_way, aes(x = known_age_ka, y = pct_diff, colour = method)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_point(alpha = 0.6, size = 1.3) +
  scale_x_log10() +
  facet_wrap(~method) +
  labs(x = "Known (independently dated) age (ka)",
       y = "Estimated − known age, as % of known age",
       title = "Age differences from known ages, all three approaches",
       subtitle = "Dashed line: zero difference; same 128 samples as above",
       colour = NULL) +
  theme_bw() +
  theme(legend.position = "none")
```

All three approaches scatter tightly around the 1:1 line, as expected
since all three are fit to (or, for the constant-parameter Bayesian
approach, taken directly from) this same calibration set. The spreadsheet
and constant-parameter Bayesian approach are nearly indistinguishable,
consistent with Stage 2b's finding that they share the same calibration
curves and combine the two acids' evidence in an algebraically equivalent
way. The estimated-parameter Bayesian approach (B) shows the same overall
pattern but with a small systematic offset from its shifted `a_Glu`/`e_Glu`
constants (Stage 1); its 66% interval coverage is the most informative
diagnostic of the three, since, unlike the median-based percent
differences, it directly checks whether each method's stated uncertainty
is calibrated against the known ages, rather than just whether the point
estimate is close.

### Which approach fits best overall?

```{r plot_three_way_density, fig.height=4}
ggplot(three_way, aes(x = pct_diff, fill = method, colour = method)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey40") +
  geom_density(alpha = 0.35, linewidth = 0.8) +
  labs(x = "Estimated minus known age, as % of known age", y = "Density",
       title = "Overall fit to known ages: distribution of % differences by method",
       fill = NULL, colour = NULL) +
  theme_bw() +
  theme(legend.position = "bottom")
```

```{r rank_vs_known, echo=FALSE}
rank_table <- three_way |>
  group_by(method) |>
  summarise(
    bias_pct             = median(pct_diff),
    median_abs_pct_diff  = median(abs(pct_diff)),
    rmse_log10_age        = sqrt(mean((log10(est_age_ka) - log10(known_age_ka))^2)),
    coverage66_pct         = 100 * mean(in_interval66),
    .groups = "drop"
  ) |>
  arrange(rmse_log10_age)

knitr::kable(rank_table, digits = 2,
             col.names = c("Method", "Bias (median signed % diff)",
                           "Median abs. % diff", "RMSE (log10 age)",
                           "66% interval coverage (%)"),
             caption = "Ranked by RMSE in log10(age), best fit first")
```

By every metric in this table, the **spreadsheet (GLS)** and **Bayesian,
constant-parameter (A)** approaches are effectively tied for best fit to
the known ages, both markedly better than the Bayesian, estimated-parameter
(B) approach: lower RMSE in log10(age), near-zero median bias (no
systematic over- or under-ageing), and 66% interval coverage close to the
nominal 66%. That is expected rather than a coincidence, since both use
the manuscript's published OLS calibration curves; Stage 2b already showed
these two are numerically almost equivalent once the curve is fixed. If a
single answer is required, the spreadsheet and Approach-A Bayesian
inversion are the best-fitting and are interchangeable in practice, so the
choice between them comes down to workflow (spreadsheet vs. R) rather than
accuracy.

Approach B fits noticeably worse (larger RMSE, a non-zero bias, and
usually the lowest, or least well-calibrated, 66% coverage), not because
Bayesian inversion is inferior as a combination method, but because its
*calibration curves* are shifted relative to the published OLS fit, for
the reason diagnosed in Stage 1 (jointly estimating `rho` alongside both
curves' shapes moves the joint likelihood's optimum away from the
per-acid OLS solution). This reinforces the Stage 2b conclusion: Approach
B is a legitimate diagnostic on the joint-fitting procedure, but not
presently a drop-in replacement calibration if the goal is to match the
independently known ages as closely as possible.

## Why Approach A agrees with GLS, and what Approach B changes

For a flat prior and Gaussian log-age residuals, the Bayesian posterior
mode and variance reduce algebraically to the GLS point estimate and
combined variance *given the same calibration curve*. Approach A supplies
exactly that shared curve, so it recovers the GLS combination almost
exactly — confirming the GLS formulation is a valid, deployable equivalent
of the Bayesian inversion. Approach B breaks that equivalence at the
curve-fitting stage, before the inversion step is ever reached: jointly
estimating the residual correlation alongside both curves' shapes yields a
different (but equally legitimate, in a strict likelihood sense) maximum
than fitting each curve's OLS regression independently and estimating
`rho` afterward, as the manuscript's spreadsheet does.

One simplification applies to both approaches: the Bayesian likelihood
here uses only the calibration-scale residual variance (`sigma_Asp`,
`sigma_Glu`), not the additional per-sample analytical-measurement-
uncertainty term the GLS calculator adds via each sample's replicate SE.
The manuscript finds that term contributes only a small amount to the
total variance on average, so this simplification has little practical
effect on the comparison above, though it can matter more for samples with
few replicates, especially young samples, where the GLS calculator's
explicit propagation of replicate uncertainty is an advantage over this
simplified Bayesian implementation.

## Approach A vs. Approach B: which to use, and when

**Approach A (adopt the manuscript's published calibration constants)**

- *Pros:* guarantees the age-inversion comparison is testing exactly one
  thing — whether a Bayesian combination of two acids' evidence reproduces
  a GLS combination of the same evidence, given the same curves. Requires
  no MCMC calibration fit (faster, nothing to tune or check for
  convergence) and cannot inherit any curve-fitting idiosyncrasy. This is
  the appropriate choice for the manuscript's section 4.1 validation
  claim specifically.
- *Cons:* it is not a fully independent, end-to-end check — it takes the
  spreadsheet's own per-acid OLS curves as ground truth rather than
  testing whether an independent statistical procedure would arrive at
  similar curves in the first place.

**Approach B (independent joint Bayesian calibration refit)**

- *Pros:* a genuinely independent, end-to-end pipeline from raw D/L
  measurements to final age, with no numbers imported from the spreadsheet
  at all. Naturally propagates calibration uncertainty into the age
  posterior (in principle — not implemented in this simplified inversion),
  and can accommodate a shared residual correlation directly in the
  likelihood rather than as an auxiliary post-hoc estimate.
- *Cons:* its point estimates diverge non-trivially from the published
  calculator (several percent median, and up to several times that for
  individual samples) for a subtle but explainable reason — jointly
  estimating the residual correlation alongside the curve shapes changes
  where the joint likelihood is maximized, relative to fitting each curve
  independently. That divergence would need to be resolved (e.g. by fixing
  `rho` post-hoc as the manuscript does, rather than estimating it jointly,
  or by using per-acid priors informative enough to anchor each curve near
  its independent OLS fit) before Approach B could be presented as an
  equally valid *replacement* calibration, rather than as a diagnostic on
  the joint-fitting procedure itself.
